An efficient composition of bidirectional programs by tupling and lazy updates

### 2019-12-03 paper/2.minbigul.tex
tiny changes in formulas: keywords --> normal text, redundant vars --> underline

        \noindent $\putbx{\casebx{cond_{sv}}{cond_{s}}{bx_1}{bx_2}}{s}{v} =\\
            \tab \text{if} \ {cond_{sv} \ s \ v} \\
            \tab \text{then} \ s' \Leftarrow \putbx{bx_1}{s}{v} \\
            \tab \text{else} \ s' \Leftarrow \putbx{bx_2}{s}{v} \\
            \tab \text{fi} \ cond_{s} \ s'; \ \text{return} \ s'$

@@

        \noindent $\getbx{\casebx{cond_{sv}}{cond_{s}}{bx_1}{bx_2}}{s} = \\
            \tab \text{if} \ {cond_{s} \ s} \\
            \tab \text{then} \ v' \Leftarrow \getbx{bx_1}{s} \\ 
            \tab \text{else} \ v' \Leftarrow \getbx{bx_2}{s} \\ 
            \tab \text{fi} \ {cond_{sv} \ s \ v'}; \ \text{return} \ v'$

@@

\smallvspace
$\tab \text{if } E_1 \text{ then } X_1 \text{ else } X_2 \text{ fi } E_2; S\\
\tabs{2} \ \equiv \text{if }E_1 = true \text{ then } \{ X_1; \text{ if } E_2 = true \text{ then } S \text{ else } undefined \}\\
    \tabs{4} \text{else } \{ X_2; \text{ if } E_2 = false \text{ then } S \text{ else } undefined \}$
\smallvspace

>>>>

        \noindent $\putbx{\casebx{cond_{sv}}{cond_{s}}{bx_1}{bx_2}}{s}{v} =\\
            \tab \textnormal{if} \ {cond_{sv} \ s \ v} \\
            \tab \textnormal{then} \ s' \Leftarrow \putbx{bx_1}{s}{v} \\
            \tab \textnormal{else} \ s' \Leftarrow \putbx{bx_2}{s}{v} \\
            \tab \textnormal{fi} \ cond_{s} \ s'; \ \textnormal{return} \ s'$

@@

        \noindent $\getbx{\casebx{cond_{sv}}{cond_{s}}{bx_1}{bx_2}}{s} = \\
            \tab \textnormal{if} \ {cond_{s} \ s} \\
            \tab \textnormal{then} \ v' \Leftarrow \getbx{bx_1}{s} \\ 
            \tab \textnormal{else} \ v' \Leftarrow \getbx{bx_2}{s} \\ 
            \tab \textnormal{fi} \ {cond_{sv} \ s \ v'}; \ \textnormal{return} \ v'$

@@

\smallvspace
$\tab \text{if } E_1 \text{ then } X_1 \text{ else } X_2 \text{ fi } E_2; S\\
\tabs{2} \ \equiv \text{if }E_1 = \textit{true} \text{ then } \{ X_1; \text{ if } E_2 = true \text{ then } S \text{ else } \textit{undefined}\}\\
    \tabs{4} \text{else } \{ X_2; \text{ if } E_2 = \textit{false} \text{ then } S \text{ else } \textit{undefined} \}$
\smallvspace

### 2019-12-03 paper/4.xpg.tex
tiny changes in formulas: keywords --> normal text, redundant vars --> underline

    \noindent $\tabs{2} \pg{Skip \ h}{s}{v} \overset{1}{=} (\sif{h \ s = v}{s}{undefined}, h \ s) \\
        \tab \qtab \overset{2}{=} \sif{h \ s = v}{(s, h \ s)}{undefined} \\
        \tab \qtab \overset{3}{=} \sif{h \ s = v}{(s, v)}{undefined}$

@@

    \noindent $\tabs{2} \pg{bx_1 \circ bx_2}{s}{v} \\
    \tabs{3} \overset{1}{=} (\putbx{bx_1}{s}{(\putbx{bx_2}{(\getbx{bx_1}{s})}{v})}, \getbx{bx_2}{(\getbx{bx_1}{s})}) \\
    \tabs{3} \overset{2}{=} v_1 \Leftarrow \getbx{bx_1}{s}; \qtabs{3} \ \, \overset{3}{=} (s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy};\\
        \tab \qtab (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v}; \qtabs{2} (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v};\\
        \tab \qtab (s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}; \qtabs{2} (s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2};\\
            \qtab \tabs{2} (s_3, v_2) \qtabs{6} (s_3, v_2)\\
    \tabs{3} \overset{4}{=} (s_1, v_1) \Leftarrow \pg{bx_1}{s}{construct\_dummy \ s}; \\
        \tab \qtab (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v}; \\
        \tab \qtab (s_3, v_3) \Leftarrow \pg{bx_1}{s_1}{s_2}; \\
            \qtab \tabs{2} (s_3, v_2)$

@@

    \noindent $\cpg{Skip \ h}{ks}{kv}{s}{v} = \text{ if } h \ s = v \text{ then } (ks, kv, s, v) \text{ else } undefined$

@@

    \noindent $\cpg{Case \ cond_{sv} \ cond_{s} \ bx_1 \ bx_2}{ks}{kv}{s}{v} =\\
        \tab \text{if} \ cond_{sv} \ s \ v \ \&\& \ cond_{s} \ s\\
        \tab \text{then} \ (ks, kv, s', v') \Leftarrow \cpg{bx_1}{ks}{kv}{s}{v}\\
        \tab \text{else} \ (ks, kv, s', v') \Leftarrow \cpg{bx_2}{ks}{kv}{s}{v}\\
        \tab \text{fi} \ cond_{s} \ s' \ \&\& \ cond_{sv} \ s \ v'; \ \text{return} \ (ks, kv, s', v')$

    \noindent $\cpg{bx_1 \circ bx_2}{ks}{kv}{s}{v} =\\
        \tab (ks_1, kv_1, s_1, v_1) \Leftarrow \cpg{bx_1}{ks}{id}{s}{construct\_dummy \ s};\\
        \tab (ks_2, kv_2, s_2, v_2) \Leftarrow \cpg{bx_2}{kv_1}{kv}{v_1}{v};\\
            \qtab (ks_1 \circ ks_2, kv_2,  ks_1 \ s_2, v_2)$
\end{definition}

@@

        Note that, in $\cpginline{bx_1 \circ bx_2}$, $s_1$ is redundant because this evaluated variable is not used in the later steps.
        In the next session, we will optimize this redundancy.

@@

    \noindent $\kpg{Skip \ h}{ks}{kv}{ks'}{kv'}{s}{v} =\\
        \tab s \Leftarrow ks' \ s; \quad v \Leftarrow kv' \ v; \quad ks' \Leftarrow id; \quad kv' \Leftarrow id;\\
        \tab \text{if} \ h \ s = v \ \text{then} \ (ks, kv, ks', kv', s, v) \ \text{else } undefined$

@@

    \noindent $\kpg{Case \ cond_{sv} \ cond_{s} \ bx_1 \ bx_2}{ks}{kv}{ks'}{kv'}{s}{v} =\\
        \tab s \Leftarrow ks' \ s; \quad v \Leftarrow kv' \ v; \quad ks' \Leftarrow id ; \quad kv' \Leftarrow id;\\
        \tab \text{if} \ cond_{sv} \ s \ v \ \&\& \ cond_{s} \ s\\
        \tab \text{then} \ (ks, kv, ks', kv', s', v') \Leftarrow \kpg{bx_1}{ks}{kv}{ks'}{kv'}{s}{v}\\
        \tab \text{else} \ (ks, kv, ks', kv', s', v') \Leftarrow \kpg{bx_2}{ks}{kv}{ks'}{kv'}{s}{v}\\
        \tab \text{fi} \ cond_s \ (ks' \ s') \ \&\& \ cond_{sv} \ s \ (kv' \ v'); \ \text{return} \ (ks, kv, ks', kv', s', v')$

    \noindent $\kpg{bx_1 \circ bx_2}{ks}{kv}{ks'}{kv'}{s}{v} =\\
        \tab (ks_1, kv_1, ks_1', kv_1', s_1, v_1) \Leftarrow \kpg{bx_1}{ks}{id}{ks'}{id}{s}{construct\_dummy \ s};\\
        \tab (ks_2, kv_2, ks_2', kv_2', s_2, v_2) \Leftarrow \kpg{bx_2}{kv_1}{kv}{kv_1'}{kv'}{v_1}{v};\\
        \qtab (ks_1 \circ ks_2, kv_2, ks_1 \circ ks_2', kv_2', s_2, v_2)$
\end{definition}

In $kpg$, basically, functions for the updates will be kept in $ks'$ and $kv'$.
$\kpginline{Skip}$ and $\kpginline{Case}$ are special cases, because they require the actual values for evaluation. Therefore we evaluate the values by application of $ks' \ s$ and $kv' \ v$ and update $ks'$ and $kv'$ with the identity functions.
%Except $\kpginline{Skip}$ and $\kpginline{Case}$ where $s$ and $v$ hold actual values, functions for the computation will be kept in $ks'$ and $kv'$.
When evaluating $\kpginline{bx_1 \circ bx_2}$, $ks_1'$ and $s_1$ in the result of the first assignment are still redundant but they are not evaluated.

Additionally we did two optimizations in $kpg$. First is in $\kpginline{bx_1 \times bx_2}$. The evaluation of $ks'$ and $kv'$ will be done independently in two assignments using $\kpginline{bx_1}$ and $\kpginline{bx_2}$. There may be same computations in $fst \circ ks'$ and $snd \circ ks'$ as well as $fst \circ kv'$ and $snd \circ kv'$.
To remove duplicate evaluations, we evaluate actual values in $s$ and $v$ before calling $\kpginline{bx_1}$.
%It is possible to evaluate actual values in $s$ and $v$ before calling $\kpginline{bx_1}$ to remove the redundancy like that. One more thing,
Second is in $\kpginline{Case}$. We need to evaluate $ks' \ s'$ and $kv' \ v'$ to check the $fi$ condition before returning the 6-tuple. Such evaluations can be done lazily to make programs run faster. We use the above small optimizations in our implementation.

@@

    \noindent $\xpg{bx}{s}{v} = \text{match } bx \text{ with }\\
    \tab \vert \ bx_1 \circ bx_2 \to \\
        \tabs{2} (ks_1,kv_1, ks_1', kv_1', s_1, v_1) \Leftarrow \kpg{bx_1}{\lambda \_.s}{id}{id}{id}{s}{construct\_dummy \ s};\\
        \tabs{2} (s_2, v_2) \Leftarrow \xpg{bx_2}{kv_1' \ v_1}{v};\\
        \tabs{3} (ks_1 \ s_2, v_2)\\
    \tab \vert \ \_ \to \text{similar to } pg$
\end{definition}

>>>>

    \noindent $\tabs{2} \pg{Skip \ h}{s}{v} \overset{1}{=} (\sif{h \ s = v}{s}{\textit{undefined}}, h \ s) \\
        \tab \qtab \overset{2}{=} \sif{h \ s = v}{(s, h \ s)}{\textit{undefined}} \\
        \tab \qtab \overset{3}{=} \sif{h \ s = v}{(s, v)}{\textit{undefined}}$

@@

    \noindent $\tabs{2} \pg{bx_1 \circ bx_2}{s}{v} \\
    \tabs{3} \overset{1}{=} (\putbx{bx_1}{s}{(\putbx{bx_2}{(\getbx{bx_1}{s})}{v})}, \getbx{bx_2}{(\getbx{bx_1}{s})}) \\
    \tabs{3} \overset{2}{=} v_1 \Leftarrow \getbx{bx_1}{s}; \qtabs{3} \ \, \overset{3}{=} (s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy};\\
        \tabs{2} \qtab (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v}; \qtabs{2} (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v};\\
        \tabs{2} \qtab (s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}; \qtabs{2} (s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2};\\
            \qtab \tabs{3} (s_3, v_2) \qtabs{6} (s_3, v_2)\\
    \tabs{3} \overset{4}{=} (s_1, v_1) \Leftarrow \pg{bx_1}{s}{construct\_dummy \ s}; \\
        \tabs{2} \qtab (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v}; \\
        \tabs{2} \qtab (s_3, v_3) \Leftarrow \pg{bx_1}{s_1}{s_2}; \\
            \qtab \tabs{3} (s_3, v_2)$

@@

    \noindent $\cpg{Skip \ h}{ks}{kv}{s}{v} = \textnormal{ if } h \ s = v \text{ then } (ks, kv, s, v) \textnormal{ else } \textit{undefined}$

@@

    \noindent $\cpg{Case \ cond_{sv} \ cond_{s} \ bx_1 \ bx_2}{ks}{kv}{s}{v} =\\
        \tab \textnormal{if} \ cond_{sv} \ s \ v \ \&\& \ cond_{s} \ s\\
        \tab \textnormal{then} \ (ks, kv, s', v') \Leftarrow \cpg{bx_1}{ks}{kv}{s}{v}\\
        \tab \textnormal{else} \ (ks, kv, s', v') \Leftarrow \cpg{bx_2}{ks}{kv}{s}{v}\\
        \tab \textnormal{fi} \ cond_{s} \ s' \ \&\& \ cond_{sv} \ s \ v'; \ \textnormal{return} \ (ks, kv, s', v')$

    \noindent $\cpg{bx_1 \circ bx_2}{ks}{kv}{s}{v} =\\
        \tab (ks_1, kv_1, \underline{s_1}, v_1) \Leftarrow \cpg{bx_1}{ks}{id}{s}{construct\_dummy \ s};\\
        \tab (ks_2, kv_2, s_2, v_2) \Leftarrow \cpg{bx_2}{kv_1}{kv}{v_1}{v};\\
            \qtab (ks_1 \circ ks_2, kv_2,  ks_1 \ s_2, v_2)$
\end{definition}

@@

Note that, in $\cpginline{bx_1 \circ bx_2}$, $\underline{s_1}$ is redundant because this evaluated variable is not used in the later steps.
In the next session, we will optimize this redundancy.

@@

    \noindent $\kpg{Skip \ h}{ks}{kv}{ks'}{kv'}{s}{v} =\\
        \tab s \Leftarrow ks' \ s; \quad v \Leftarrow kv' \ v; \quad ks' \Leftarrow id; \quad kv' \Leftarrow id;\\
        \tab \textnormal{if} \ h \ s = v \ \textnormal{then} \ (ks, kv, ks', kv', s, v) \ \textnormal{else } \text{undefined}$

@@

    \noindent $\kpg{Case \ cond_{sv} \ cond_{s} \ bx_1 \ bx_2}{ks}{kv}{ks'}{kv'}{s}{v} =\\
        \tab s \Leftarrow ks' \ s; \quad v \Leftarrow kv' \ v; \quad ks' \Leftarrow id ; \quad kv' \Leftarrow id;\\
        \tab \textnormal{if} \ cond_{sv} \ s \ v \ \&\& \ cond_{s} \ s\\
        \tab \textnormal{then} \ (ks, kv, ks', kv', s', v') \Leftarrow \kpg{bx_1}{ks}{kv}{ks'}{kv'}{s}{v}\\
        \tab \textnormal{else} \ (ks, kv, ks', kv', s', v') \Leftarrow \kpg{bx_2}{ks}{kv}{ks'}{kv'}{s}{v}\\
        \tab \underline{\textnormal{fi}} \ cond_s \ (ks' \ s') \ \&\& \ cond_{sv} \ s \ (kv' \ v'); \ \textnormal{return} \ (ks, kv, ks', kv', s', v')$

    \noindent $\kpg{bx_1 \circ bx_2}{ks}{kv}{ks'}{kv'}{s}{v} =\\
        \tab (ks_1, kv_1, \underline{ks_1'}, kv_1', \underline{s_1}, v_1) \Leftarrow \kpg{bx_1}{ks}{id}{ks'}{id}{s}{construct\_dummy \ s};\\
        \tab (ks_2, kv_2, ks_2', kv_2', s_2, v_2) \Leftarrow \kpg{bx_2}{kv_1}{kv}{kv_1'}{kv'}{v_1}{v};\\
        \qtab (ks_1 \circ ks_2, kv_2, ks_1 \circ ks_2', kv_2', s_2, v_2)$
\end{definition}

In $kpg$, basically, functions for the updates will be kept in $ks'$ and $kv'$.
$\kpginline{Skip}$ and $\kpginline{Case}$ are special cases, because they require the actual values for evaluation. Therefore we evaluate the values by application of $ks' \ s$ and $kv' \ v$ and update $ks'$ and $kv'$ with the identity functions.
%Except $\kpginline{Skip}$ and $\kpginline{Case}$ where $s$ and $v$ hold actual values, functions for the computation will be kept in $ks'$ and $kv'$.
When evaluating $\kpginline{bx_1 \circ bx_2}$, $\underline{ks_1'}$ and $\underline{s_1}$ in the result of the first assignment are still redundant but they are not evaluated.

Additionally we did two optimizations in $kpg$. First is in $\kpginline{bx_1 \times bx_2}$. The evaluation of $ks'$ and $kv'$ will be done independently in two assignments using $\kpginline{bx_1}$ and $\kpginline{bx_2}$. There may be same computations in $fst \circ ks'$ and $snd \circ ks'$ as well as $fst \circ kv'$ and $snd \circ kv'$.
To remove duplicate evaluations, we evaluate actual values in $s$ and $v$ before calling $\kpginline{bx_1}$.
%It is possible to evaluate actual values in $s$ and $v$ before calling $\kpginline{bx_1}$ to remove the redundancy like that. One more thing,
Second is in $\kpginline{Case}$. We need to evaluate $ks' \ s'$ and $kv' \ v'$ to check the \underline{fi} condition before returning the 6-tuple. Such evaluations can be done lazily to make programs run faster. We use the above small optimizations in our implementation.

@@

    \noindent $\xpg{bx}{s}{v} = \textnormal{match } bx \textnormal{ with }\\
    \tab \vert \ bx_1 \circ bx_2 \to \\
        \tabs{2} (ks_1,kv_1, ks_1', kv_1', s_1, v_1) \Leftarrow \kpg{bx_1}{\lambda \_.s}{id}{id}{id}{s}{construct\_dummy \ s};\\
        \tabs{2} (s_2, v_2) \Leftarrow \xpg{bx_2}{kv_1' \ v_1}{v};\\
        \tabs{3} (ks_1 \ s_2, v_2)\\
    \tab \vert \ \_ \to \text{similar to } pg$
\end{definition}
